---
title: "Giuliani PCSR immune cleaning"
author: "Michelle Byrne"
date: "June 19, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
knitr::opts_knit$set(root.dir = '/home/michelle/Dropbox/academic/collaboration/giuliani_immune')
```

## All processing and cleaning for immune markers in Nicole Giuliani's PCSR project

Calculate intra-assay CVs (from optical density) from the sample (duplicate) CVs:

```{r intra_cvs}
intraassays <- read_csv("intraassays.csv")
View(intraassays)
crp_pl1_cv <- mean(intraassays$crp_pl1)
crp_pl2_cv <- mean(intraassays$crp_pl2)
crp_cv <- (crp_pl1_cv + crp_pl2_cv)/2
il1b_pl1_cv <- mean(intraassays$il1b_pl1)
il1b_pl2_cv <- mean(intraassays$il1b_pl2)
il1b_cv <- (il1b_pl1_cv + il1b_pl2_cv)/2
siga_pl1_cv <- mean(intraassays$siga_pl1)
siga_pl2_cv <- mean(intraassays$siga_pl2)
siga_cv <- (siga_pl1_cv + siga_pl2_cv)/2
il6_pl1_cv <- mean(intraassays$il6_pl1)
il6_pl2_cv <- mean(intraassays$il6_pl2)
il6_cv <- (il6_pl1_cv + il6_pl2_cv)/2
```

Calculate inter-assay CVs (from optical density) from the high and low controls:

```{r inter_cvs}
interassays <- read_csv("interassays.csv")
View(interassays)
# Calculate mean for controls by plate:
interassays$plate_mean <- rowMeans(interassays[,c(4:5)])

# CRP:
# Mean of high controls across plates:
crp_high_mean <- mean(data.matrix(interassays[interassays$marker=="crp"&interassays$control=="high","plate_mean"]))
# SD of high controls across plates:
crp_high_sd <- sd(data.matrix(interassays[interassays$marker=="crp"&interassays$control=="high","plate_mean"]))
# % CV high across plates:
crp_high_cv <- (crp_high_sd/crp_high_mean)*100
# Mean of low controls across plates:
crp_low_mean <- mean(data.matrix(interassays[interassays$marker=="crp"&interassays$control=="low","plate_mean"]))
# SD of low controls across plates:
crp_low_sd <- sd(data.matrix(interassays[interassays$marker=="crp"&interassays$control=="low","plate_mean"]))
# % CV low across plates:
crp_low_cv <- (crp_low_sd/crp_low_mean)*100
# AVERAGE % CV:
crp_inter_cv <- (crp_low_cv + crp_high_cv)/2

# IL-1B:
# Mean of high controls across plates:
il1b_high_mean <- mean(data.matrix(interassays[interassays$marker=="il1b"&interassays$control=="high","plate_mean"]))
# SD of high controls across plates:
il1b_high_sd <- sd(data.matrix(interassays[interassays$marker=="il1b"&interassays$control=="high","plate_mean"]))
# % CV high across plates:
il1b_high_cv <- (il1b_high_sd/il1b_high_mean)*100
# Mean of low controls across plates:
il1b_low_mean <- mean(data.matrix(interassays[interassays$marker=="il1b"&interassays$control=="low","plate_mean"]))
# SD of low controls across plates:
il1b_low_sd <- sd(data.matrix(interassays[interassays$marker=="il1b"&interassays$control=="low","plate_mean"]))
# % CV low across plates:
il1b_low_cv <- (il1b_low_sd/il1b_low_mean)*100
# AVERAGE % CV:
il1b_inter_cv <- (il1b_low_cv + il1b_high_cv)/2

# SIgA:
# Mean of high controls across plates:
siga_high_mean <- mean(data.matrix(interassays[interassays$marker=="siga"&interassays$control=="high","plate_mean"]))
# SD of high controls across plates:
siga_high_sd <- sd(data.matrix(interassays[interassays$marker=="siga"&interassays$control=="high","plate_mean"]))
# % CV high across plates:
siga_high_cv <- (siga_high_sd/siga_high_mean)*100
# Mean of low controls across plates:
siga_low_mean <- mean(data.matrix(interassays[interassays$marker=="siga"&interassays$control=="low","plate_mean"]))
# SD of low controls across plates:
siga_low_sd <- sd(data.matrix(interassays[interassays$marker=="siga"&interassays$control=="low","plate_mean"]))
# % CV low across plates:
siga_low_cv <- (siga_low_sd/siga_low_mean)*100
# AVERAGE % CV:
siga_inter_cv <- (siga_low_cv + siga_high_cv)/2

# IL-6:
# Mean of high controls across plates:
il6_high_mean <- mean(data.matrix(interassays[interassays$marker=="il6"&interassays$control=="high","plate_mean"]))
# SD of high controls across plates:
il6_high_sd <- sd(data.matrix(interassays[interassays$marker=="il6"&interassays$control=="high","plate_mean"]))
# % CV high across plates:
il6_high_cv <- (il6_high_sd/il6_high_mean)*100
# Mean of low controls across plates:
il6_low_mean <- mean(data.matrix(interassays[interassays$marker=="il6"&interassays$control=="low","plate_mean"]))
# SD of low controls across plates:
il6_low_sd <- sd(data.matrix(interassays[interassays$marker=="il6"&interassays$control=="low","plate_mean"]))
# % CV low across plates:
il6_low_cv <- (il6_low_sd/il6_low_mean)*100
# AVERAGE % CV:
il6_inter_cv <- (il6_low_cv + il6_high_cv)/2

```

Load data and deal with out of range values (too high or low for assay to detect)

```{r}
pcsr_immune <- read_csv("raw_concentrations.csv")
View(pcsr_immune)
increment = 0.001
oor_right = "OOR>"
oor_left = "OOR<"
num_oor_right_crp <- length(na.omit(pcsr_immune$crp_raw [pcsr_immune$crp_raw == oor_right]))
num_oor_right_il1b <- length(na.omit(pcsr_immune$il1b_raw [pcsr_immune$il1b_raw == oor_right]))
num_oor_right_siga <- length(na.omit(pcsr_immune$siga_raw [pcsr_immune$siga_raw == oor_right]))
num_oor_right_il6 <- length(na.omit(pcsr_immune$il6_raw [pcsr_immune$il6_raw == oor_right]))

num_oor_left_crp <- length(na.omit(pcsr_immune$crp_raw [pcsr_immune$crp_raw == oor_left]))
num_oor_left_il1b <- length(na.omit(pcsr_immune$il1b_raw [pcsr_immune$il1b_raw == oor_left]))
num_oor_left_siga <- length(na.omit(pcsr_immune$siga_raw [pcsr_immune$siga_raw == oor_left]))
num_oor_left_il6 <- length(na.omit(pcsr_immune$il6_raw [pcsr_immune$il6_raw == oor_left]))

# Report these in your manuscript:
num_oor_right_crp 
num_oor_right_il1b 
num_oor_right_siga 
num_oor_right_il6 
num_oor_left_crp 
num_oor_left_il1b 
num_oor_left_siga 
num_oor_left_il6 

# Now replace (winsorize) the OOR samples

# First identify what the max/min are outside of the OOR values (and force them to be numeric, because if they had OOR in them they wouldn't be):

crp_num <- pcsr_immune[!(pcsr_immune$crp_raw==oor_right),] 
crp_num$crp_raw <- as.numeric(crp_num$crp_raw)
max_crp <- max(na.omit(crp_num$crp_raw))
min_crp <- min(na.omit(crp_num$crp_raw))

il1b_num <- pcsr_immune[!(pcsr_immune$il1b_raw==oor_right),]
il1b_num$il1b_raw <- as.numeric(il1b_num$il1b_raw)
max_il1b <- max(na.omit(il1b_num$il1b_raw))
min_il1b <- min(na.omit(il1b_num$il1b_raw))

siga_num <- pcsr_immune[!(pcsr_immune$siga_raw==oor_right),]
siga_num$siga_raw <- as.numeric(siga_num$siga_raw)
max_siga <- max(na.omit(siga_num$siga_raw))
min_siga <- min(na.omit(siga_num$siga_raw))

il6_num <- pcsr_immune[!(pcsr_immune$il6_raw==oor_right),]
il6_num$il6_raw <- as.numeric(il6_num$il6_raw)
max_il6 <- max(na.omit(il6_num$il6_raw))
min_il6 <- min(na.omit(il6_num$il6_raw))

# If the immune marker is OOR> then replace with max plus increment, OOR< then replace with min minus increment
pcsr_immune$crp_raw <- ifelse (pcsr_immune$crp_raw==oor_right,max_crp + increment,pcsr_immune$crp_raw)
pcsr_immune$il1b_raw <- ifelse (pcsr_immune$il1b_raw==oor_right,max_il1b + increment,pcsr_immune$il1b_raw)
pcsr_immune$siga_raw <- ifelse (pcsr_immune$siga_raw==oor_right,max_siga + increment,pcsr_immune$siga_raw)
pcsr_immune$il6_raw <- ifelse (pcsr_immune$il6_raw==oor_right,max_il6 + increment,pcsr_immune$il6_raw)

pcsr_immune$crp_raw <- ifelse (pcsr_immune$crp_raw==oor_left,min_crp - increment,pcsr_immune$crp_raw)
pcsr_immune$il1b_raw <- ifelse (pcsr_immune$il1b_raw==oor_left,min_il1b - increment,pcsr_immune$il1b_raw)
pcsr_immune$siga_raw <- ifelse (pcsr_immune$siga_raw==oor_left,min_siga - increment,pcsr_immune$siga_raw)
pcsr_immune$il6_raw <- ifelse (pcsr_immune$il6_raw==oor_left,min_il6 - increment,pcsr_immune$il6_raw)

# Then force everything to be numeric for further analyses
pcsr_immune$crp_raw <- as.numeric(pcsr_immune$crp_raw)
pcsr_immune$il1b_raw <- as.numeric(pcsr_immune$il1b_raw)
pcsr_immune$siga_raw <- as.numeric(pcsr_immune$siga_raw)
pcsr_immune$il6_raw <- as.numeric(pcsr_immune$il6_raw)
str(pcsr_immune)

```

Correct your data for flow rate
#### Flow rate = *(Volume of saliva)/(time sample took in sec)*
#### Corrected concentration = *(observed concentration) x (Flow rate)*
##### CRP is not flow rate dependent
##### SIgA *is* flow rate dependent
##### IL-1 beta does not appear to be flow rate dependent in individuals with IL-1 beta levels in the normal range
##### IL-6 - no evidence (yet) that it is flow rate dependent but who knows

```{r}
vol_ml = 2.5
pcsr_immune$fr <- vol_ml / pcsr_immune$duration_sec
pcsr_immune$siga_frc <- pcsr_immune$siga_raw * pcsr_immune$fr 
```


Check normality, outliers, and skew stats in the raw concentrations
```{r check_norm}

# Histograms for immune markers to visualize normality
hist(pcsr_immune$crp_raw)
hist(pcsr_immune$il1b_raw)
hist(pcsr_immune$siga_frc)
hist(pcsr_immune$il6_raw)

#Boxplots to visualize outliers
library(ggplot2)

ggplot(pcsr_immune, aes(x = "", y = crp_raw)) +   
  geom_boxplot() +
  ylab("CRP (ng/ml)") +
  ggtitle("PCSR CRP") +
  geom_smooth(method='lm', color="black")
ggplot(pcsr_immune, aes(x = "", y = il1b_raw)) +   
  geom_boxplot() +
  ylab("IL1-B (ng/ml)") +
  ggtitle("PCSR IL1-B") +
  geom_smooth(method='lm', color="black")
ggplot(pcsr_immune, aes(x = "", y = siga_frc)) +   
  geom_boxplot() +
  ylab("SIgA (ng/ml)") +
  ggtitle("PCSR SIgA") +
  geom_smooth(method='lm', color="black")
ggplot(pcsr_immune, aes(x = "", y = il6_raw)) +   
  geom_boxplot() +
  ylab("IL-6 (ng/ml)") +
  ggtitle("PCSR IL-6") +
  geom_smooth(method='lm', color="black")

#  Calculate and report normality statistics (skew and kurtosis)
# Kurtosis and Skew should be -2/+2 (West, et al. 1995)
library(psych)
crp_summary <- describe(pcsr_immune$crp_raw)
crp_summary
il1b_summary <- describe(pcsr_immune$il1b_raw)
il1b_summary
siga_summary <- describe(pcsr_immune$siga_frc)
siga_summary
il6_summary <- describe(pcsr_immune$il6_raw)
il6_summary

```

If necessary, transform data

```{r}
 
# if skew or kurtosis is < -2 or > 2 (or in the case of CRP, visual inspection shows it's not normal), then calculate the natural log of that variable
#Note: although it is common to do a natural log or square root transformation, these transfromations did not result in acceptable skew/kurtosis stats. Therefore, we chose to use the transform Tukey funciton, which uses iterative Shapiro-Wilk tests, and finds the lambda values that maximizes the W statistic from those tests. (e.g., find power transformation that makes the data fit the normal distribution as closely as possible). See info on rcompanion for more.
library(rcompanion)

if (crp_summary$skew > 2 | crp_summary$skew < -2 | crp_summary$kurtosis > 2 | crp_summary$kurtosis < -2) {
  pcsr_immune$t_crp <- transformTukey(pcsr_immune$crp_raw)
}

# transform CRP anyway because visual inspection shows it's not normal
pcsr_immune$t_crp <- transformTukey(pcsr_immune$crp_raw)

if (il1b_summary$skew > 2 | il1b_summary$skew < -2 | il1b_summary$kurtosis > 2 | il1b_summary$kurtosis < -2) {
  pcsr_immune$t_il1b <- transformTukey(pcsr_immune$il1b_raw)
}
if (siga_summary$skew > 2 | siga_summary$skew < -2 | siga_summary$kurtosis > 2 | siga_summary$kurtosis < -2) {
  pcsr_immune$t_siga <- transformTukey(pcsr_immune$siga_frc)
}
if (il6_summary$skew > 2 | il6_summary$skew < -2 | il6_summary$kurtosis > 2 | il6_summary$kurtosis < -2) {
  pcsr_immune$t_il6 <- transformTukey(pcsr_immune$il6_raw)
}

# Check everything post transformation for any transformed variables 
final_crp_summary <- describe(pcsr_immune$t_crp) 
final_il1b_summary <- describe(pcsr_immune$t_il1b)
final_siga_summary <- describe(pcsr_immune$t_siga)
final_il6_summary <- describe(pcsr_immune$t_il6)

hist(pcsr_immune$t_crp)
hist(pcsr_immune$t_il1b)
hist(pcsr_immune$t_siga)
hist(pcsr_immune$t_il6)

ggplot(pcsr_immune, aes(x = "", y = t_crp)) +   
  geom_boxplot() +
  ylab("Tukey (CRP)") +
  ggtitle("PCSR CRP (transformed)") +
  geom_smooth(method='lm', color="black")
ggplot(pcsr_immune, aes(x = "", y = t_il1b)) +   
  geom_boxplot() +
  ylab("Tukey (IL1-B)") +
  ggtitle("PCSR IL1-B (transformed)") +
  geom_smooth(method='lm', color="black")
ggplot(pcsr_immune, aes(x = "", y = t_siga)) +   
  geom_boxplot() +
  ylab("Tukey (SIgA)") +
  ggtitle("PCSR SIgA (transformed)") +
  geom_smooth(method='lm', color="black")
ggplot(pcsr_immune, aes(x = "", y = t_il6)) +   
  geom_boxplot() +
  ylab("Tukey (IL-6)") +
  ggtitle("PCSR IL-6 (transformed)") +
  geom_smooth(method='lm', color="black")

```

Winsorize any remaining outliers that are +/- 3 SD (if any)

```{r}
crp_up_limit <- (final_crp_summary$mean + 3*(final_crp_summary$sd))
crp_up_limit
crp_lo_limit <- (final_crp_summary$mean - 3*(final_crp_summary$sd))
crp_lo_limit
il1b_up_limit <- (final_il1b_summary$mean + 3*(final_il1b_summary$sd))
il1b_up_limit
il1b_lo_limit <- (final_il1b_summary$mean - 3*(final_il1b_summary$sd))
il1b_lo_limit
siga_up_limit <- (final_siga_summary$mean + 3*(final_siga_summary$sd))
siga_up_limit
siga_lo_limit <- (final_siga_summary$mean - 3*(final_siga_summary$sd))
siga_lo_limit
il6_up_limit <- (final_il6_summary$mean + 3*(final_il6_summary$sd))
il6_up_limit
il6_lo_limit <- (final_il6_summary$mean - 3*(final_il6_summary$sd))
il6_lo_limit

# First check if there are any outliers (if not, no need to winsorize). Also, report these numbers. (in this case there were none, so there is no more code for winsorizing here)
crp_num_right_outliers <- length(na.omit(pcsr_immune$t_crp [pcsr_immune$t_crp > crp_up_limit])) 
crp_num_right_outliers
crp_num_left_outliers <- length(na.omit(pcsr_immune$t_crp [pcsr_immune$t_crp < crp_lo_limit])) 
crp_num_left_outliers 
il1b_num_right_outliers <- length(na.omit(pcsr_immune$t_il1b [pcsr_immune$t_il1b > il1b_up_limit]))
il1b_num_right_outliers
il1b_num_left_outliers <- length(na.omit(pcsr_immune$t_il1b [pcsr_immune$t_il1b < il1b_lo_limit])) 
il1b_num_left_outliers
siga_num_right_outliers <- length(na.omit(pcsr_immune$t_siga [pcsr_immune$t_siga > siga_up_limit]))
siga_num_right_outliers 
siga_num_left_outliers <- length(na.omit(pcsr_immune$t_siga [pcsr_immune$t_siga < siga_lo_limit])) 
siga_num_left_outliers
il6_num_right_outliers <- length(na.omit(pcsr_immune$t_il6 [pcsr_immune$t_il6 > il6_up_limit])) 
il6_num_right_outliers
il6_num_left_outliers <- length(na.omit(pcsr_immune$t_il6 [pcsr_immune$t_il6 < il6_lo_limit])) 
il6_num_left_outliers

```
In this data, there are no outliers, but here is some code for how you would do that.
```{r}
#library(DescTools)
#pcsr_immune$siga_w <- Winsorize(pcsr_immune$t_siga, minval = siga_lo_limit, maxval = siga_up_limit, na.rm = TRUE)
#pcsr_immune$il6_w <- Winsorize(pcsr_immune$t_il6, minval = il6_lo_limit, maxval = il6_up_limit, na.rm = TRUE)
```


Output final cleaned data
```{r}
pcsr_immune$crp_cleaned <- pcsr_immune$t_crp
pcsr_immune$il1b_cleaned <- pcsr_immune$t_il1b
pcsr_immune$siga_cleaned <- pcsr_immune$t_siga
pcsr_immune$il6_cleaned <- pcsr_immune$t_il6

write.csv(pcsr_immune, file="immune_cleaned.csv")
```

